---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
    theme: united
---

## Preparing the environment

The libraries that are used throughout this R Markdown file are installed if necessary and then loaded by use of the following dependency function.

```{r results="hide", warning=F, message=F}
depend <- function (s) {
	if (s %in% rownames(installed.packages()) == FALSE) {
		install.packages(s, repos="http://cran.rstudio.com/", dependencies=TRUE)
	}
	library(s, character.only=TRUE)
}
```

## Loading and preprocessing the data

We'll read the raw data set directly, taking care to preserve headers and format *NA* values. Date's can be tricky when imported, so we'll force the *date* column to be interpretted as a date.

```{r, cache=T}
dat <- read.table(unz("activity.zip", "activity.csv"), na.strings = "NA", header=T, quote="\"", sep=",")
dat$date <- as.Date(dat$date)
```

Inspecting the data shows well formated column names as well as good data for the *date* and *interval* columns.

```{r, cache=T}
head(dat)
```
Inspecting the end of the table shows that the *interval* column continues to count through the entire period. It does not start again each day.

```{r, cache=T}
tail(dat)
```

A quick sum shows that there is good data further down in the *steps* column as well.

```{r}
sum(dat$steps,na.rm=TRUE)
```

## What is mean total number of steps taken per day?

### Libraries

We'll be using the *ggplot2* library for all plots. The *scales* library helps us with breaks and graph limiting, and *ggthemes* brings us a wonderful collection of colors.

Tidying the data will be done by with the *dplyr* library and *magrittr* piping.

```{r results="hide", warning=F, message=F}
depend("ggplot2")
depend("scales")
depend("ggthemes")
depend("dplyr")
```

### Steps per day

To calculate the number of steps per day, we group all of our data by each day and take the sum of each group. The result table can be used to draw a plot for each day.

```{r}
steps_per_day <- dat %>%
    select( steps, date ) %>%
    group_by( date) %>%
    summarize ( steps=sum(steps) )
```

```{r warning=F, fig.width=10, fig.height=6}
ggplot(steps_per_day, aes(date,steps), na.rm = TRUE) +
    geom_point(colour="white", size=1) +
    theme_solarized(light=FALSE) +
    scale_colour_solarized("red") +
    labs(x="Date", y="Steps", title="Number of Steps per Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5)) +
    scale_x_date(breaks = "1 day", labels=date_format("%d %B %y"))
```

### Histogram of steps per day

In addition to the literal number of steps per day, we can also look at a histogram of step intervals to identify the shape of the data and the likelihood of outliers.

```{r warning=F, fig.width=10, fig.height=6}
ggplot(steps_per_day, aes(steps)) +
    geom_histogram(binwidth=1000, colour="deepskyblue1", fill="white") +
    theme_solarized(light=FALSE) +
    scale_colour_solarized("red") +
    labs(x="Steps", y="Frequency", title="Histogram of Steps per Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank())
```

### Mean and Median

With the *steps_per_day* table, calculating the mean and median values per day is a straight forward operation.

```{r}
mean(steps_per_day$steps, na.rm=TRUE)
median(steps_per_day$steps, na.rm=TRUE)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

