---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
    theme: united
---

## Preparing the environment

The libraries that are used throughout this R Markdown file are installed if
necessary and then loaded by use of the following dependency function.

```{r results="hide", warning=F, message=F}
depend <- function (s) {
	if (s %in% rownames(installed.packages()) == FALSE) {
		install.packages(s, repos="http://cran.rstudio.com/", dependencies=TRUE)
	}
	library(s, character.only=TRUE)
}
```

## Loading and preprocessing the data

We'll read the raw data set directly, taking care to preserve headers and
format *NA* values. Date's can be tricky when imported, so we'll force the
*date* column to be interpretted as a date.

```{r, cache=T}
dat <- read.table(unz("activity.zip", "activity.csv"),
                  na.strings = "NA", header=T, quote="\"", sep=",")
dat$date <- as.Date(dat$date)
```

Inspecting the data shows well formated column names as well as good data for
the *date* and *interval* columns.

```{r, cache=T}
head(dat)
```

A quick sum shows that there is good data further down in the *steps* column as
well.

```{r}
sum(dat$steps,na.rm=TRUE)
```

## What is mean total number of steps taken per day?

### Libraries

We'll be using the *ggplot2* library for all plots. The *scales* library helps
us with breaks and graph limiting, and *ggthemes* brings us a wonderful
collection of colors.

Tidying the data will be done by with the *dplyr* library and *magrittr* piping.

```{r results="hide", warning=F, message=F}
depend("ggplot2")
depend("scales")
depend("ggthemes")
depend("dplyr")
```

### Steps per day

To calculate the number of steps per day, we group all of our data by each day
and take the sum of each group. The result table can be used to draw a plot for
each day.

```{r}
steps_per_day <- dat %>%
    select(steps, date) %>%
    group_by(date) %>%
    summarize(steps=sum(steps))
```

```{r warning=F, fig.width=10, fig.height=6}
ggplot(steps_per_day, aes(date,steps), na.rm = TRUE) +
    geom_point(size=2, color="#099DD9") +
    theme_solarized() +
    scale_colour_solarized("red") +
    labs(x="Date", y="Steps", title="Number of Steps per Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5)) +
    scale_x_date(breaks = "1 day", labels=date_format("%d %B %y"))
```

### Histogram of steps per day

In addition to the literal number of steps per day, we can also look at a
histogram of step intervals to identify the shape of the data and the
likelihood of outliers.

```{r warning=F, fig.width=10, fig.height=6}
ggplot(steps_per_day, aes(steps)) +
    geom_histogram(binwidth=500, color = "black", fill = "#099DD9") +
    theme_solarized() +
    labs(x="Steps", y="Frequency", title="Histogram of Steps per Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank())
```

### Mean and Median

With the *steps_per_day* table, calculating the mean and median values per day
is a straight forward operation.

```{r}
mean(steps_per_day$steps, na.rm=TRUE)
median(steps_per_day$steps, na.rm=TRUE)
```

## What is the average daily activity pattern?

### Libraries

To better access date-time information from the tables, we will use the
*lubridate* library. To facilitate the tidy step, *stringr* will give the
ability to pad a numeric string, helpful in the time transformation below.

```{r results="hide", warning=F, message=F}
depend("lubridate")
depend("stringr")
```

### Time Series

The *interval* column is really nothing more than a badly formatted time
column. By grouping by that column and taking the average of the steps, we can
plot a time series.

```{r}
time_series <- dat %>%
    mutate(interval=parse_date_time(str_pad(interval, 4, pad="0"), "%H%M")) %>%
    group_by(interval) %>%
    summarize(steps=mean(steps, na.rm=TRUE))
head(time_series)
```

```{r warning=F, fig.width=10, fig.height=6}
ggplot(time_series, aes(interval,steps)) +
    geom_line(color="#099DD9") +
    theme_solarized() +
    scale_colour_solarized("red") +
    labs(x="Time", y="Steps", title="Average Number of Steps During Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5)) +
    scale_x_datetime(breaks = "60 min", labels=date_format("%H %M"))
```

### Highest average 5-minute interval

Looking at the graph, it is easy to estimate the highest peak falling slightly
after 8:30am. We can get a more accurate picture by looking for the row with
the maximum step value, then formatting the time interval to be human readible.

```{r}
max_row <- time_series[which.max(time_series$steps),]
max_interval <- max_row$interval
paste(hour(max_interval),minute(max_interval),sep=":")
```

8:35am matches the graph expectations.

## Imputing missing values

### How many NA values

A number of the rows in this data set contain missing values. Looking row-wise
at the entire data set, we can count how many have *NA* values and how many do
not.

```{r}
na_dat <- dat[!complete.cases(dat),]
complete_dat <- dat[complete.cases(dat),]
nrow(na_dat)
nrow(complete_dat)
```

### Filling in the gaps

We can attempt to infer the values of the NA positions by leveraging the mean
step count for that 5-minute interval and inserting that value in place of the
NA.

- First we will format the interval column to match the datetime format used in
  our lookup table, *time_series*.
- We identify which rows contain missing *steps* information.
- We identify the *interval* value for each of these rows.
- We use this *interval* value to look up the replacement *steps* value from
  *time_series*.

```{r}
dat_fill <- dat %>%
    mutate(interval=parse_date_time(str_pad(interval, 4, pad="0"), "%H%M"))

na.steps <- which(is.na(dat_fill$steps))
na.interval <- dat_fill$interval[na.steps]
dat_fill$steps[na.steps] <- time_series$steps[match(na.interval, time_series$interval)]

head(dat_fill)
```

### Histogram of changes

Producing a histogram of the new table, *dat_fill*, shows that by filling in
the missing NA values with the average for that time period across all days,
the shape and distribution of the data remains fairly constant, however, the
overall frequency has increased for each interval.

   is the impact of imputing missing data on the estimates of the total daily
   number of steps?

```{r}
steps_per_day_fill <- dat_fill %>%
    select(steps, date) %>%
    group_by(date) %>%
    summarize(steps=sum(steps))
```

```{r warning=F, fig.width=10, fig.height=6}
ggplot(steps_per_day_fill, aes(steps)) +
    geom_histogram(binwidth=500, color = "black", fill = "#099DD9") +
    theme_solarized() +
    labs(x="Steps", y="Frequency", title="Histogram of Steps per Day (Filled NAs)") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank())
```

### Mean and Median

In calculating the mean & median values of the *steps_per_day_fill* table,
there is a small change in the median, but no visible change in the mean. This
suggests that the missing NA values were evenly distrubuted so that by filling
them we have not shifted anything.

```{r}
mean(steps_per_day_fill$steps)
median(steps_per_day_fill$steps)
```

Looking at the total number of steps from the original data table versus our
filled data table reveals just how many new steps have been interpolated.

```{r}

total_steps <- sum(dat$steps,na.rm=TRUE)
total_steps

total_steps_fill <- sum(dat_fill$steps)
total_steps_fill

total_steps_fill - total_steps
```

## Are there differences in activity patterns between weekdays and weekends?

We create a new factor variable in the dataset, *week*, with two levels –
“weekday” and “weekend” indicating whether a given date is a weekday or weekend
day.

```{r}
dat_fill <- dat_fill %>%
    mutate ( week = ifelse( wday(date, label = TRUE, abbr = FALSE) %in% 
                           c("Saturday", "Sunday"), "weekend", "weekday"))

head(dat_fill)
```

Finally, we can plot a time series of our filled data set, breaking the weekday
or weekend factors by color. Overlaying the plots shows the obvious differences
in steps throughout the day for a weekend vs weekday.

```{r warning=F, fig.width=10, fig.height=6}
time_series_fill <- dat_fill %>%
    group_by(interval, week) %>%
    summarize(steps=mean(steps, na.rm=TRUE))

ggplot(time_series_fill, aes(interval, steps, col=week) ) +
    geom_line( ) +
    theme_solarized() +
    scale_colour_solarized("red") +
    labs(x="Time", y="Steps", title="Average Number of Steps During Day (Filled NAs)") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5)) +
    scale_x_datetime(breaks = "60 min", labels=date_format("%H %M"))
```
