---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
    theme: united
---

## Preparing the environment

The libraries that are used throughout this R Markdown file are installed if
necessary and then loaded by use of the following dependency function.

```{r results="hide", warning=F, message=F}
depend <- function (s) {
	if (s %in% rownames(installed.packages()) == FALSE) {
		install.packages(s, repos="http://cran.rstudio.com/", dependencies=TRUE)
	}
	library(s, character.only=TRUE)
}
```

## Loading and preprocessing the data

We'll read the raw data set directly, taking care to preserve headers and
format *NA* values. Date's can be tricky when imported, so we'll force the
*date* column to be interpretted as a date.

```{r, cache=T}
dat <- read.table(unz("activity.zip", "activity.csv"),
                  na.strings = "NA", header=T, quote="\"", sep=",")
dat$date <- as.Date(dat$date)
```

Inspecting the data shows well formated column names as well as good data for
the *date* and *interval* columns.

```{r, cache=T}
head(dat)
```

A quick sum shows that there is good data further down in the *steps* column as
well.

```{r}
sum(dat$steps,na.rm=TRUE)
```

## What is mean total number of steps taken per day?

### Libraries

We'll be using the *ggplot2* library for all plots. The *scales* library helps
us with breaks and graph limiting, and *ggthemes* brings us a wonderful
collection of colors.

Tidying the data will be done by with the *dplyr* library and *magrittr* piping.

```{r results="hide", warning=F, message=F}
depend("ggplot2")
depend("scales")
depend("ggthemes")
depend("dplyr")
```

### Steps per day

To calculate the number of steps per day, we group all of our data by each day
and take the sum of each group. The result table can be used to draw a plot for
each day.

```{r}
steps_per_day <- dat %>%
    select(steps, date) %>%
    group_by(date) %>%
    summarize(steps=sum(steps))
```

```{r warning=F, fig.width=10, fig.height=6}
ggplot(steps_per_day, aes(date,steps), na.rm = TRUE) +
    geom_point(size=2, color="#099DD9") +
    theme_solarized() +
    scale_colour_solarized("red") +
    labs(x="Date", y="Steps", title="Number of Steps per Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5)) +
    scale_x_date(breaks = "1 day", labels=date_format("%d %B %y"))
```

### Histogram of steps per day

In addition to the literal number of steps per day, we can also look at a
histogram of step intervals to identify the shape of the data and the
likelihood of outliers.

```{r warning=F, fig.width=10, fig.height=6}
ggplot(steps_per_day, aes(steps)) +
    geom_histogram(binwidth=500, color = "black", fill = "#099DD9") +
    theme_solarized() +
    labs(x="Steps", y="Frequency", title="Histogram of Steps per Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank())
```

### Mean and Median

With the *steps_per_day* table, calculating the mean and median values per day
is a straight forward operation.

```{r}
mean(steps_per_day$steps, na.rm=TRUE)
median(steps_per_day$steps, na.rm=TRUE)
```

## What is the average daily activity pattern?

### Libraries

To better access date-time information from the tables, we will use the
*lubridate* library. To facilitate the tidy step, *stringr* will give the
ability to pad a numeric string, helpful in the time transformation below.

```{r results="hide", warning=F, message=F}
depend("lubridate")
depend("stringr")
```

### Time Series

The *interval* column is really nothing more than a badly formatted time
column. By grouping by that column and taking the average of the steps, we can
plot a time series.

```{r}
time_series <- dat %>%
    mutate(interval=parse_date_time(str_pad(interval, 4, pad="0"), "%H%M")) %>%
    group_by(interval) %>%
    summarize(steps=mean(steps, na.rm=TRUE))
```
```{r warning=F, fig.width=10, fig.height=6}
ggplot(time_series, aes(interval,steps)) +
    geom_line(color="#099DD9") +
    theme_solarized() +
    scale_colour_solarized("red") +
    labs(x="Time", y="Steps", title="Average Number of Steps During Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5)) +
    scale_x_datetime(breaks = "60 min", labels=date_format("%H %M"))
```

### Highest average 5-minute interval

Looking at the graph, it is easy to estimate the highest peak falling slightly
after 8:30am. We can get a more accurate picture by looking for the row with
the maximum step value, then formatting the time interval to be human readible.

```{r}
max_row <- time_series[which.max(time_series$steps),]
max_interval <- max_row$interval
paste(hour(max_interval),minute(max_interval),sep=":")
```

8:35am matches the graph expectations.

## Imputing missing values

### How many NA values

A number of the rows in this data set contain missing values. Looking row-wise
at the entire data set, we can count how many have *NA* values and how many do
not.

```{r}
na_dat <- dat[!complete.cases(dat),]
complete_dat <- dat[complete.cases(dat),]
nrow(na_dat)
nrow(complete_dat)
```

### Filling in the gaps

We can attempt to infer the values of the NA positions by leveraging the mean
step count for that 5-minute interval and inserting that value in place of the
NA.

```{r}
dat_fill <- dat %>%
    mutate(interval=parse_date_time(str_pad(interval, 4, pad="0"), "%H%M")) %>%
    rowwise() %>%
    mutate(steps_filled = ifelse(is.na(steps), 0, steps))

head(dat_fill)
```
3. Create a new dataset that is equal to the original dataset but with the
   missing data filled in.

4. Make a histogram of the total number of steps taken each day and Calculate
   and report the mean and median total number of steps taken per day. Do these
   values differ from the estimates from the first part of the assignment? What
   is the impact of imputing missing data on the estimates of the total daily
   number of steps?


```{r}
#high_res_dat <- dat %>%
#    mutate(interval=str_pad(interval, 4, pad="0")) %>%
#    mutate(datetime=parse_date_time(paste(date, interval), "%Y-%m-%d %H%M")) %>%
#    select(steps, datetime)
#
#head(high_res_dat)
```



## Are there differences in activity patterns between weekdays and weekends?

