---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
    theme: united
---

## Preparing the environment

The libraries that are used throughout this R Markdown file are installed if necessary and then loaded by use of the following dependency function.

```{r results="hide", warning=F, message=F}
depend <- function (s) {
	if (s %in% rownames(installed.packages()) == FALSE) {
		install.packages(s, repos="http://cran.rstudio.com/", dependencies=TRUE)
	}
	library(s, character.only=TRUE)
}
```

## Loading and preprocessing the data

```{r results="hide", warning=F, message=F}
depend("dplyr")
depend("lubridate")
```

### Loading the data set

We'll read the raw data set directly, taking care to preserve headers and format NA values.

```{r, cache=T}
dat <- read.table(unz("activity.zip", "activity.csv"), na.strings = "NA", header=T, quote="\"", sep=",")
dat$date <- as.Date(dat$date)
```

Inspecting the data shows well formated column names as well as good data for the date and interval columns.

```{r, cache=T}
head(dat)
```
Inspecting the end of the table shows that the interval column continues to count through the entire period. It does not start again each day.

```{r, cache=T}
tail(dat)
```

A quick sum shows that there is good data further down in the steps column as well.

```{r}
sum(dat$steps,na.rm=TRUE)
```

## What is mean total number of steps taken per day?

### Libraries

We'll be using the ggplot2 library for all plots. Scales library helps us with breaks and graph limiting, and ggthemes brings us a wonderful collection of colors.

```{r results="hide", warning=F, message=F}
depend("ggplot2")
depend("scales")
depend("ggthemes")
```

```{r fig.width=10, fig.height=6}
ggplot(dat, aes(x=date, y=steps)) +
    geom_histogram() +
    theme_solarized(light=FALSE) +
    scale_colour_solarized("red") +
    labs(x="Date", y="Steps", title="Histogram of Steps per Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5)) +
    scale_x_date(breaks = "1 day", labels=date_format("%d %B %y"))
```


## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?

