---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
    theme: united
---

## Preparing the environment

The libraries that are used throughout this R Markdown file are installed if
necessary and then loaded by use of the following dependency function.

```{r results="hide", warning=F, message=F}
depend <- function (s) {
	if (s %in% rownames(installed.packages()) == FALSE) {
		install.packages(s, repos="http://cran.rstudio.com/", dependencies=TRUE)
	}
	library(s, character.only=TRUE)
}
```

## Loading and preprocessing the data

We'll read the raw data set directly, taking care to preserve headers and
format *NA* values. Date's can be tricky when imported, so we'll force the
*date* column to be interpretted as a date.

```{r, cache=T}
dat <- read.table(unz("activity.zip", "activity.csv"),
                  na.strings = "NA", header=T, quote="\"", sep=",")
dat$date <- as.Date(dat$date)
```

Inspecting the data shows well formated column names as well as good data for
the *date* and *interval* columns.

```{r, cache=T}
head(dat)
```

A quick sum shows that there is good data further down in the *steps* column as
well.

```{r}
sum(dat$steps,na.rm=TRUE)
```

## What is mean total number of steps taken per day?

### Libraries

We'll be using the *ggplot2* library for all plots. The *scales* library helps
us with breaks and graph limiting, and *ggthemes* brings us a wonderful
collection of colors.

Tidying the data will be done by with the *dplyr* library and *magrittr* piping.

```{r results="hide", warning=F, message=F}
depend("ggplot2")
depend("scales")
depend("ggthemes")
depend("dplyr")
```

### Steps per day

To calculate the number of steps per day, we group all of our data by each day
and take the sum of each group. The result table can be used to draw a plot for
each day.

```{r}
steps_per_day <- dat %>%
    select(steps, date) %>%
    group_by(date) %>%
    summarize(steps=sum(steps))
```

```{r warning=F, fig.width=10, fig.height=6}
ggplot(steps_per_day, aes(date,steps), na.rm = TRUE) +
    geom_point(colour="white", size=1) +
    theme_solarized(light=FALSE) +
    scale_colour_solarized("red") +
    labs(x="Date", y="Steps", title="Number of Steps per Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5)) +
    scale_x_date(breaks = "1 day", labels=date_format("%d %B %y"))
```

### Histogram of steps per day

In addition to the literal number of steps per day, we can also look at a
histogram of step intervals to identify the shape of the data and the
likelihood of outliers.

```{r warning=F, fig.width=10, fig.height=6}
ggplot(steps_per_day, aes(steps)) +
    geom_histogram(binwidth=1000, colour="deepskyblue1", fill="white") +
    theme_solarized(light=FALSE) +
    scale_colour_solarized("red") +
    labs(x="Steps", y="Frequency", title="Histogram of Steps per Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank())
```

### Mean and Median

With the *steps_per_day* table, calculating the mean and median values per day
is a straight forward operation.

```{r}
mean(steps_per_day$steps, na.rm=TRUE)
median(steps_per_day$steps, na.rm=TRUE)
```

## What is the average daily activity pattern?

### Libraries

To better access date-time information from the tables, we will use the
*lubridate* library. To facilitate the tidy step, *stringr* will give the
ability to pad a numeric string, helpful in the time transformation below.

```{r results="hide", warning=F, message=F}
depend("lubridate")
depend("stringr")
```

### Time Series

The *interval* column is really nothing more than a badly formatted time
column. By grouping by that column and taking the average of the steps, we can
plot a time series.

```{r}
time_series <- dat %>%
    mutate(interval=parse_date_time(str_pad(interval, 4, pad="0"), "%H%M")) %>%
    group_by(interval) %>%
    summarize(steps=mean(steps, na.rm=TRUE))
```
```{r warning=F, fig.width=10, fig.height=6}
ggplot(time_series, aes(interval,steps)) +
    geom_line(colour="white") +
    theme_solarized(light=FALSE) +
    scale_colour_solarized("red") +
    labs(x="Date", y="Steps", title="Number of Steps During Day") +
    theme(plot.title=element_text(size=15, face="bold", vjust=2)) +
    theme(legend.title=element_blank()) +
    theme(axis.text.x=element_text(angle=50, size=5, vjust=0.5)) +
    scale_x_datetime(breaks = "60 min", labels=date_format("%H %M"))
```

## Imputing missing values

```{r}
#high_res_dat <- dat %>%
#    mutate(interval=str_pad(interval, 4, pad="0")) %>%
#    mutate(datetime=parse_date_time(paste(date, interval), "%Y-%m-%d %H%M")) %>%
#    select(steps, datetime)
#
#head(high_res_dat)
```



## Are there differences in activity patterns between weekdays and weekends?

